---
title: "School diversity hearing"
author: "Nick Solomon and James Subudhi"
date: '`r councildown::pretty_date()`'
output: 
  councildown::council_html: default
  councildown::council_pdf:
    latex_engine: xelatex
compact-title: true
fontsize: 11pt
subparagraph: yes
---

```{r include = FALSE}
library(ggplot2)
library(councildown)
library(scales)

extrafont::loadfonts()
knitr::knit_hooks$set(embed = hook_pdfembed)
knitr::opts_chunk$set(echo = FALSE, embed = TRUE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = here::here())
```

```{r load-data}
source("code/00_load_data.R")
```


# City Overview

This section contains facts about all schools and students in New York City that were included in the 2017 committee report. All data is for the 2017-18 school year.

## Race and Ethnicity

```{r}
citywide <- read_excel("data/original_data/demographicsnapshot201314to201718public_final.xlsx", sheet = "Citywide") %>% 
  clean_names()

cols <- c("#706AE0", "#CB5871",  "#82C91E", "#16AC9E", "#F59F00")

citywide %>% 
  filter(year == "2017-18") %>% 
  select(number_asian,
         number_black,
         number_hispanic,
         number_multiple_race_categories_not_represented,
         number_white) %>%
  gather("race", "number") %>%
  mutate(prop = number/sum(number)) %>%
  mutate(race = str_remove(race, "number_") %>%
           str_replace_all("_", " ") %>%
           str_to_sentence() %>%
           str_wrap(width = 30) %>% 
           reorder(-number)) %>% 
  ggplot(aes(race, prop)) +
  geom_col(aes(fill = race), show.legend = FALSE) +
  geom_text(aes(label = paste0(number(number,big.mark = ","), " (", percent(prop) ,")")),
            family = "Times New Roman",
            vjust = -.5) +
  scale_fill_manual(values = cols) +
  scale_y_continuous(labels = percent_format(accuracy = 1), expand = expand_scale(mult = .05, add = c(0, .01))) +
  theme_nycc(print = TRUE) +
  labs(title = "Demographic breakdown of New York City Public Schools",
       subtitle = "For the 2017-18 school year",
       x = "Race/ethnicity",
       y = "Percent of students",
       caption = "Source: DOE Demographic Snapshot")
```



```{r}
num_black_hispanic_less_10_white <- school_dems %>% 
  filter(year == "2017-18", percent_white < .1, !str_detect(dbn, "^84")) %>% 
  summarize(number_black = sum(number_black),
            number_hispanic = sum(number_hispanic)) %>% 
  mutate(total = number_black + number_hispanic) %>% 
  pull(total)

num_black_hispanic <- school_dems %>% 
  filter(year == "2017-18", !str_detect(dbn, "^84")) %>% 
  summarize(number_black = sum(number_black),
            number_hispanic = sum(number_hispanic)) %>% 
  mutate(total = number_black + number_hispanic) %>% 
  pull(total)


num_white_more_50_white <- school_dems %>% 
  filter(year == "2017-18", percent_white > .5, !str_detect(dbn, "^84")) %>% 
  summarize(number_white = sum(number_white)) %>% 
  pull(number_white)

num_white <- school_dems %>% 
  filter(year == "2017-18", !str_detect(dbn, "^84")) %>% 
  summarize(number_white = sum(number_white)) %>% 
  pull(number_white)

```

- `r percent(num_black_hispanic_less_10_white/num_black_hispanic)` of black and Hispanic students attend a school that is less than 10% white students.
- `r percent(num_white_more_50_white/num_white)` of white students attend a school with more than 50% white students

```{r}
more_75_poverty_race <- school_dems %>% 
  filter(year == "2017-18", !str_detect(dbn, "^84"), percent_poverty > .75) %>% 
  summarize_at(vars(starts_with("number_")), sum) %>% 
  select(number_asian,
         number_black,
         number_hispanic,
         number_multiple_race_categories_not_represented,
         number_white)

num_race <- school_dems %>% 
  filter(year == "2017-18", !str_detect(dbn, "^84")) %>% 
  summarize_at(vars(starts_with("number_")), sum) %>% 
  select(number_asian,
         number_black,
         number_hispanic,
         number_multiple_race_categories_not_represented,
         number_white)

 

dat <- (more_75_poverty_race/num_race) %>% 
  gather("race", "prop") %>% 
  mutate(race = str_remove(race, "number_") %>%
           str_replace_all("_", " ") %>%
           str_to_sentence() %>%
           str_wrap(width = 30) %>% 
           reorder(-prop))
cols <- c("#706AE0", "#CB5871",  "#82C91E", "#F59F00", "#16AC9E")
names(cols) <- levels(dat$race)
  
ggplot(dat, aes(race, prop)) +
  geom_col(aes(fill = race), show.legend = FALSE) +
  scale_fill_manual(values = cols) + 
   geom_text(aes(label = percent(prop)),
            family = "Times New Roman",
            vjust = -.5) +
  scale_y_continuous(labels = percent_format(accuracy = 1), expand = expand_scale(mult = .05, add = c(0, .01))) +
  labs(title = "Percentage of students attending a school where\nmore than 75% of students experience poverty",
       x = "Race/ethnicity",
       y = "Percent of students",
       caption = "Source: DOE Demographic Snapshot") + 
  theme_nycc(print = TRUE)
```

## Poverty

```{r include = FALSE}
city_pov <- citywide %>% 
  filter(year == "2017-18") %>% 
  pull(percent_poverty)

city_eni <- citywide %>% 
  filter(year == "2017-18") %>% 
  pull(economic_need_index)
```


- Citywide, `r percent(city_pov)` of students experience poverty. The citywide Economic Need Index is `r percent(city_eni)`.

## ELL Students

```{r}
num_more_20_ell <- school_dems %>% 
  filter(year == "2017-18", !str_detect(dbn, "^84"), percent_english_language_learners > .2) %>% 
  nrow()
num_less_10_ell <- school_dems %>% 
  filter(year == "2017-18", !str_detect(dbn, "^84"), percent_english_language_learners < .1) %>% 
  nrow()

num_schools <- school_dems %>% 
  filter(year == "2017-18", !str_detect(dbn, "^84")) %>% 
  nrow()
```

- `r percent(num_more_20_ell/num_schools)` of schools have a population of more than 20% ELL students
- `r percent(num_less_10_ell/num_schools)` of schools have a population of less than 10% ELL students

## Students With Disabilities

- `r citywide %>% filter(year == "2017-18") %>% pull(percent_students_with_disabilities) %>% percent()` of students in New York City are students with disabilities.

```{r}
num_less_10_dis <- school_dems %>% 
  filter(year == "2017-18", !str_detect(dbn, "^84"), percent_students_with_disabilities < .1) %>% 
  nrow()
```

- `r percent(num_less_10_dis/num_schools)` of schools have a population that is less than 10% students with disabilities.

## Students in temporary housing

```{r}
diversity_report_k8 <- read_excel("data/original_data/17-18_report.xlsx", sheet = "Grades K-8 School") %>% 
  clean_names()

diversity_report_912 <- read_excel("data/original_data/17-18_report.xlsx", sheet = "Grades 9-12 School") %>% 
  clean_names()

all_temp_housing <- bind_rows(diversity_report_912, diversity_report_k8) %>% 
  filter(category == "Students in temporary housing", !str_detect(dbn, "^84")) %>%
  select(dbn, temp_housing = total_enrollment) %>%
  mutate(temp_housing = parse_number(temp_housing)) %>% 
  group_by(dbn) %>% 
  summarize(temp_housing = sum(temp_housing, na.rm = TRUE))

total_enrollment <- bind_rows(diversity_report_912, diversity_report_k8) %>% 
  filter(category == "All Students", !str_detect(dbn, "^84")) %>%
  select(dbn, total_enrollment) %>%
  mutate(total_enrollment = parse_number(total_enrollment)) %>% 
  group_by(dbn) %>% 
  summarize(total_enrollment = sum(total_enrollment, na.rm = TRUE))

perc_more_15_temp_housing <- total_enrollment %>% 
  left_join(all_temp_housing, by = "dbn") %>% 
  mutate(perc_temp_housing = temp_housing/total_enrollment) %>%
  replace_na(list(perc_temp_housing = 0)) %>% 
  summarize(mean(perc_temp_housing > .15, na.rm = TRUE)) %>% 
  pull()

perc_no_temp_housing <- total_enrollment %>% 
  left_join(all_temp_housing, by = "dbn") %>% 
  mutate(perc_temp_housing = temp_housing/total_enrollment) %>% 
  summarize(mean(perc_temp_housing == 0, na.rm = TRUE)) %>% 
  pull()
```


- `r percent(perc_more_15_temp_housing)` of schools have a population of more than 15% of students in temporary housing.
- `r percent(perc_no_temp_housing)` of schools do not have any students in temporary housing.

# Specialized High School Diversity

Specialized high schools have been an important part of the conversation around school diversity and integration in New York City. The graphic below shows the racial demographics of 8 of the 9 specialized high schools. LaGuardia High School is not show here as it has auditions based admissions. Clearly, severe discrepancies exist for those given access to the highest level of academic achievement New York City has to offer.

```{r fig.asp=1}
source("code/shs_diversity.R")
shs_demo_plot
```


## Poverty in Specialized High Schools

```{r}
school_dems %>%
  filter(dbn %in% shs, year == "2017-18") %>%
  select(school_name,
        percent_poverty, economic_need_index) %>% 
  mutate_if(is.numeric, percent) %>% 
  knitr::kable(col.names = c("School", "Percent of Students Experiencing Poverty", "Economic Need Index"))
```

